---
title: "Homework 3: Multiple Linear Regression — Solutions"
subtitle: "MATH 4780 / MSSC 5780 · Fall 2026"
editor: source
---

# Statistical Computing and Data Analysis

Use the same **Table B.3 automobile gasoline mileage data** from Homework 2.

```{r}
#| label: setup
#| include: false
# Locate the course bundle when rendering from the homework folder or
# from another location within the course project.
candidates <- c("./data/regression_datasets.rda",
                "../data/regression_datasets.rda",
                "./regression_datasets.rda",
                "./slides/data/regression_datasets.rda",
                "../slides/data/regression_datasets.rda")
data_path <- candidates[file.exists(candidates)][1]
if (is.na(data_path)) {
  stop("Place regression_datasets.rda in ./data/ or adjust candidates in the setup chunk.")
}
load(data_path)
stopifnot(exists("table_B3"), nrow(table_B3) == 32,
          all(c("y", "x1") %in% names(table_B3)))
mpg_data <- table_B3

fit <- lm(y ~ x1 + x6, data = mpg_data)
fit_slr <- lm(y ~ x1, data = mpg_data)

summ_fit <- summary(fit)
n <- nrow(mpg_data)
```

## 1. Fit and interpret the multiple linear regression model

Fit

$$
Y_i=\beta_0+\beta_1x_{1i}+\beta_6x_{6i}+\varepsilon_i.
$$

```{r}
summary(fit)
coef(fit)
```

The fitted regression equation is approximately

$$
\widehat{y}
=
32.8846
-0.05315x_1
+0.9592x_6.
$$

The coefficient of engine displacement is $b_1=-0.05315$. Holding the number of carburetor barrels $x_6$ fixed, an increase of one cubic inch in engine displacement is associated with a decrease of about $0.053$ mpg in predicted gasoline mileage.

The coefficient of the number of carburetor barrels is $b_6=0.9592$. Holding engine displacement $x_1$ fixed, one additional carburetor barrel is associated with an increase of about $0.959$ mpg in predicted gasoline mileage.

The word **holding** is important: these are partial regression coefficients, so each coefficient describes the association with mileage after accounting for the other predictor in the model.

## 2. Overall significance of regression

We test

$$
H_0:\beta_1=\beta_6=0
$$

against

$$
H_1:\beta_j\neq 0 \text{ for at least one }j.
$$

As discussed in class, this is an **overall $F$ test** comparing the full model with the intercept-only null model.

```{r}
summ_fit
```

The reported overall test gives approximately

$$
F=53.6688
$$

with $2$ and $29$ degrees of freedom, and

$$
p\approx1.79\times10^{-10}.
$$

The same test can be viewed directly as a comparison of nested models:

```{r}
null_model <- lm(y ~ 1, data = mpg_data)
anova(null_model, fit)
```

Under the null model, every fitted value is $\bar y$, so its residual sum of squares is $SS_T$. The reduction in residual sum of squares from the null model to the full model is therefore $SS_R$.

The overall ANOVA quantities are approximately

| Source | df | SS | MS | $F$ |
|:--|--:|--:|--:|--:|
| Regression | 2 | 974.3095 | 487.1548 | 53.6688 |
| Residual | 29 | 263.2345 | 9.0771 | |
| Total | 31 | 1237.5441 | | |

Since the $p$-value is much smaller than $0.05$, we reject $H_0$. There is strong evidence that **at least one** of engine displacement and number of carburetor barrels contributes linearly to explaining gasoline mileage.

## 3. $R^2$ and adjusted $R^2$

For the multiple linear regression model,

```{r}
summ_fit$r.squared
summ_fit$adj.r.squared
```

we obtain

$$
R^2=0.7873,
\qquad
R_{\mathrm{Adj}}^2=0.7726.
$$

For the simple linear regression model using only $x_1$,

```{r}
summary(fit_slr)$r.squared
summary(fit_slr)$adj.r.squared
```

we obtain

$$
R^2=0.7723,
\qquad
R_{\mathrm{Adj}}^2=0.7647.
$$

Adding $x_6$ increases $R^2$ from about $0.772$ to $0.787$. Because $R^2$ cannot decrease when a predictor is added, the adjusted $R^2$ is more useful for judging whether the added variable provides enough improvement to offset the additional model complexity.

Here, adjusted $R^2$ also increases, but only modestly, from about $0.765$ to $0.773$.

## 4. Confidence interval for $\beta_1$

A 95% confidence interval for the regression coefficients can be obtained with

```{r}
confint(fit, level = 0.95)
```

For $\beta_1$, the 95% confidence interval is approximately

$$
\boxed{(-0.06570,\,-0.04060)}.
$$

Thus, after accounting for the number of carburetor barrels, we estimate that increasing engine displacement by one cubic inch is associated with a decrease of between about $0.041$ and $0.066$ mpg in the mean gasoline mileage.

Since zero is not contained in the interval, the data provide evidence that $\beta_1\neq0$ at the 5% significance level.

## 5. Marginal test for $x_6$

We test

$$
H_0:\beta_6=0
\qquad\text{versus}\qquad
H_1:\beta_6\neq0.
$$

The coefficient table is

```{r}
summary(fit)$coefficients
```

For $x_6$,

$$
t=1.4311,
\qquad
df=29,
\qquad
p=0.1631.
$$

Since $p>0.05$, we fail to reject $H_0$. After accounting for engine displacement, there is not sufficient evidence at the 5% significance level that the number of carburetor barrels has a nonzero linear effect on mean gasoline mileage.

## 6. Confidence interval for the mean response

For

$$
x_1=275
\qquad\text{and}\qquad
x_6=2,
$$

use

```{r}
new_car <- data.frame(x1 = 275, x6 = 2)

predict(
  fit,
  newdata = new_car,
  interval = "confidence",
  level = 0.95
)
```

The point estimate of the mean gasoline mileage is approximately

$$
\hat y=20.1874\text{ mpg}.
$$

The 95% confidence interval for the **mean gasoline mileage** is approximately

$$
\boxed{(18.8722,\ 21.5026)\text{ mpg}}.
$$

We are 95% confident that the mean gasoline mileage for automobiles with engine displacement 275 in$^3$ and two carburetor barrels lies within this interval, under the fitted model.

## 7. Prediction for a new automobile

Use

```{r}
predict(
  fit,
  newdata = new_car,
  interval = "prediction",
  level = 0.95
)
```

The point prediction is

$$
\boxed{20.1874\text{ mpg}},
$$

and the 95% prediction interval is approximately

$$
\boxed{(13.8867,\ 26.4881)\text{ mpg}}.
$$

This interval predicts the gasoline mileage of a single new automobile with engine displacement 275 in$^3$ and two carburetor barrels.

The prediction interval is wider than the confidence interval in Question 6 because it includes both uncertainty in estimating the mean response and the natural car-to-car variability around that mean.

## 8. Compare uncertainty with the simple linear regression model

For the simple linear regression model from Homework 2,

```{r}
new_car_slr <- data.frame(x1 = 275)

CI_slr <- predict(
  fit_slr,
  newdata = new_car_slr,
  interval = "confidence",
  level = 0.95
)

PI_slr <- predict(
  fit_slr,
  newdata = new_car_slr,
  interval = "prediction",
  level = 0.95
)

CI_slr
PI_slr
```

For the multiple linear regression model,

```{r}
CI_mlr <- predict(
  fit,
  newdata = new_car,
  interval = "confidence",
  level = 0.95
)

PI_mlr <- predict(
  fit,
  newdata = new_car,
  interval = "prediction",
  level = 0.95
)

CI_mlr
PI_mlr
```

The interval lengths are

```{r}
CI_slr[1, "upr"] - CI_slr[1, "lwr"]
CI_mlr[1, "upr"] - CI_mlr[1, "lwr"]

PI_slr[1, "upr"] - PI_slr[1, "lwr"]
PI_mlr[1, "upr"] - PI_mlr[1, "lwr"]
```

Approximately,

| Interval | SLR | MLR |
|:--|--:|--:|
| Mean-response CI length | 2.2214 | 2.6304 |
| Prediction-interval length | 12.7146 | 12.6014 |

At this particular predictor setting, adding $x_6$ does **not** reduce uncertainty in estimating the mean response: the confidence interval becomes wider.

The prediction interval becomes only slightly shorter, from about $12.715$ mpg to $12.601$ mpg. Thus, for this particular prediction, adding $x_6$ provides only a very small reduction in prediction uncertainty. This is consistent with the marginal test in Question 5, where $x_6$ was not statistically significant after accounting for $x_1$.

## 9. Verify $SS_{\mathrm{res}}$ using matrix operations

Construct the response vector and design matrix:

```{r}
y <- mpg_data$y

X <- cbind(
  1,
  mpg_data$x1,
  mpg_data$x6
)

colnames(X) <- c("Intercept", "x1", "x6")

X[1:5, ]
```

The least-squares estimator is

$$
\mathbf b
=
(\mathbf X'\mathbf X)^{-1}\mathbf X'\mathbf y.
$$

Compute it directly:

```{r}
b <- solve(t(X) %*% X) %*% t(X) %*% y
b
```

This agrees with

```{r}
coef(fit)
```

Now compute the residual sum of squares using matrix operations:

```{r}
e <- y - X %*% b

t(e) %*% e
```

and compare with

```{r}
sum(residuals(fit)^2)
```

Both calculations give approximately

$$
\boxed{SS_{\mathrm{res}}=263.2345}.
$$

Therefore,

$$
(\mathbf y-\mathbf X\mathbf b)'
(\mathbf y-\mathbf X\mathbf b)
=
\sum_{i=1}^n(y_i-\hat y_i)^2.
$$

## 10. Predictor effect plots

As discussed in class, a predictor effect plot displays the fitted relationship between the response and one predictor while accounting for the other predictors in the model.

We use the `effects` package in the same way as in the lecture notes.

```{r}
#| fig-width: 10
#| fig-height: 5

par(mfrow = c(1, 2))

plot(
  effects::predictorEffects(
    predictor = "x1",
    mod = fit
  ),
  main = expression("Predictor effect of " * x[1])
)

plot(
  effects::predictorEffects(
    predictor = "x6",
    mod = fit
  ),
  main = expression("Predictor effect of " * x[6])
)

par(mfrow = c(1, 1))
```

For $x_1$, the fitted effect slopes downward. Holding the other predictor fixed, larger engine displacement is associated with lower mean gasoline mileage. This agrees with the estimated coefficient $b_1=-0.05315$.

For $x_6$, the fitted effect slopes upward because $b_6=0.9592$. However, the confidence band is relatively wide, consistent with the marginal test in Question 5, where there was not sufficient evidence at the 5% level that $\beta_6$ differs from zero.

The important interpretation is that each plot shows the effect of one predictor **given that the other predictor is already in the model**.

## 11. Joint confidence region for $(\beta_1,\beta_6)$

The individual intervals from `confint()` are **marginal** confidence intervals. They consider one coefficient at a time and do not account for the correlation between the coefficient estimates.

First examine the estimated covariance and correlation matrices:

```{r}
V <- vcov(fit)
V
cov2cor(V)
```

To construct the 95% joint confidence region for $(\beta_1,\beta_6)$, use the same `car::confidenceEllipse()` approach used in the lecture notes.

```{r}
ci <- confint(fit)

par(mgp = c(2.8, 0.9, 0), mar = c(4, 4, 2, 1))

car::confidenceEllipse(
  fit,
  levels = 0.95,
  which.coef = c("x1", "x6"),
  xlab = expression(beta[1]),
  ylab = expression(beta[6]),
  main = expression(
    paste("95% Confidence Region for ", beta[1], " and ", beta[6])
  )
)

# Marginal 95% confidence intervals
abline(v = ci["x1", ], lty = 2)
abline(h = ci["x6", ], lty = 2)

# Estimated coefficient pair
points(
  coef(fit)["x1"],
  coef(fit)["x6"],
  pch = 19
)

# Joint null value
points(0, 0, pch = 4, cex = 1.5)
```

The ellipse is the 95% **joint confidence region** for the pair $(\beta_1,\beta_6)$. It reflects both the uncertainty in the two coefficient estimates and their correlation.

The dashed vertical and horizontal lines show the separate 95% marginal confidence intervals. The joint region is not simply the rectangle formed by those marginal intervals because $b_1$ and $b_6$ are correlated.

The point $(0,0)$ is outside the joint confidence region. This agrees with the overall $F$-test in Question 2: we reject

$$
H_0:\beta_1=\beta_6=0.
$$

This does **not** mean that both coefficients are individually significant. In Question 5, $\beta_6=0$ was still plausible in the marginal test, whereas $\beta_1=0$ was not.

# Mathematical Derivation and Reasoning

## 12. Relationship between $r^2$ and $R^2$

For simple linear regression with an intercept, define

$$
S_{xx}
=
\sum_{i=1}^n(x_i-\bar x)^2,
$$

$$
S_{yy}
=
\sum_{i=1}^n(y_i-\bar y)^2,
$$

and

$$
S_{xy}
=
\sum_{i=1}^n(x_i-\bar x)(y_i-\bar y).
$$

The sample correlation coefficient is

$$
r
=
\frac{S_{xy}}{\sqrt{S_{xx}S_{yy}}},
$$

so

$$
r^2
=
\frac{S_{xy}^2}{S_{xx}S_{yy}}.
$$

In simple linear regression,

$$
b_1=\frac{S_{xy}}{S_{xx}}.
$$

Also,

$$
\hat y_i-\bar y
=
b_1(x_i-\bar x).
$$

Therefore,

$$
\begin{aligned}
SS_R
&=
\sum_{i=1}^n(\hat y_i-\bar y)^2\\
&=
b_1^2
\sum_{i=1}^n(x_i-\bar x)^2\\
&=
b_1^2S_{xx}\\
&=
\left(\frac{S_{xy}}{S_{xx}}\right)^2S_{xx}\\
&=
\frac{S_{xy}^2}{S_{xx}}.
\end{aligned}
$$

Since

$$
SS_T=S_{yy},
$$

we have

$$
R^2
=
\frac{SS_R}{SS_T}
=
\frac{S_{xy}^2/S_{xx}}{S_{yy}}
=
\frac{S_{xy}^2}{S_{xx}S_{yy}}
=
r^2.
$$

Hence,

$$
\boxed{r^2=R^2}.
$$

## 13. Properties of the hat matrix

Let

$$
\mathbf H
=
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'.
$$

### (a) $\mathbf H$ is symmetric and idempotent

Because $\mathbf X'\mathbf X$ is symmetric,

$$
\left[(\mathbf X'\mathbf X)^{-1}\right]'
=
(\mathbf X'\mathbf X)^{-1}.
$$

Therefore,

$$
\begin{aligned}
\mathbf H'
&=
\left[
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'
\right]'\\
&=
\mathbf X
\left[(\mathbf X'\mathbf X)^{-1}\right]'
\mathbf X'\\
&=
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'\\
&=
\mathbf H.
\end{aligned}
$$

Thus, $\mathbf H$ is symmetric.

Next,

$$
\begin{aligned}
\mathbf H^2
&=
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'\\
&=
\mathbf X(\mathbf X'\mathbf X)^{-1}
(\mathbf X'\mathbf X)
(\mathbf X'\mathbf X)^{-1}\mathbf X'\\
&=
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'\\
&=
\mathbf H.
\end{aligned}
$$

Therefore, $\mathbf H$ is idempotent.

### (b) $\mathbf I-\mathbf H$ is symmetric and idempotent

Since both $\mathbf I$ and $\mathbf H$ are symmetric,

$$
(\mathbf I-\mathbf H)'
=
\mathbf I-\mathbf H.
$$

Thus, $\mathbf I-\mathbf H$ is symmetric.

Also,

$$
\begin{aligned}
(\mathbf I-\mathbf H)^2
&=
\mathbf I-2\mathbf H+\mathbf H^2\\
&=
\mathbf I-2\mathbf H+\mathbf H\\
&=
\mathbf I-\mathbf H,
\end{aligned}
$$

because $\mathbf H^2=\mathbf H$.

Hence, $\mathbf I-\mathbf H$ is also idempotent.

## 14. Trace of the hat matrix

We have

$$
\mathbf H
=
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'.
$$

Using the cyclic property of the trace,

$$
\operatorname{tr}(\mathbf A\mathbf B)
=
\operatorname{tr}(\mathbf B\mathbf A),
$$

we obtain

$$
\begin{aligned}
\operatorname{tr}(\mathbf H)
&=
\operatorname{tr}
\left[
\mathbf X(\mathbf X'\mathbf X)^{-1}\mathbf X'
\right]\\
&=
\operatorname{tr}
\left[
(\mathbf X'\mathbf X)^{-1}\mathbf X'\mathbf X
\right]\\
&=
\operatorname{tr}(\mathbf I_p)\\
&=
p.
\end{aligned}
$$

Therefore,

$$
\boxed{\operatorname{tr}(\mathbf H)=p}.
$$
