---
title: "Lab 7: Logistic Regression"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    css: lab-style.css
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

Fit and interpret a logistic regression model, work with probabilities,
log-odds, and odds ratios, assess model fit, and use fitted probabilities for
classification.

```{r}
library(ggplot2)
library(dplyr)
library(broom)
```

## 1. Fastener data

The `fastener` dataset from the 2026 logistic-regression slides contains:

- `load`: applied load,
- `m`: number of fasteners tested,
- `y`: number with the event of interest.

```{r}
fastener
```

Create the observed proportion.

```{r}
fastener <- fastener |>
  mutate(prop = y / m)

ggplot(fastener, aes(load, prop)) +
  geom_point(size = 3) +
  labs(x = "Load", y = "Observed event proportion") +
  coord_cartesian(ylim = c(0, 1))
```

::: {.question}
Why would an ordinary linear regression of `prop` on `load` be problematic?
:::

## 2. Fit a grouped-binomial logistic model

For grouped binomial data, use the numbers of successes and failures.

```{r}
fit_logit <- glm(
  cbind(y, m - y) ~ load,
  family = binomial(link = "logit"),
  data = fastener
)

summary(fit_logit)
```

The model is

\[
\log\left(\frac{\pi(x)}{1-\pi(x)}\right)
 = \beta_0 + \beta_1 x.
\]

## 3. Interpret the coefficients

```{r}
broom::tidy(fit_logit, conf.int = TRUE)
```

Exponentiating a coefficient converts a change in log-odds into an **odds
ratio**.

```{r}
exp(coef(fit_logit))
exp(confint.default(fit_logit))
```

::: {.question}
Interpret \(\exp(\hat\beta_1)\): how are the odds expected to change for a
one-unit increase in load?
:::

## 4. Fitted probabilities

```{r}
grid <- data.frame(
  load = seq(min(fastener$load), max(fastener$load), length.out = 200)
)

grid$prob <- predict(fit_logit, newdata = grid, type = "response")

ggplot(fastener, aes(load, prop)) +
  geom_point(size = 3) +
  geom_line(data = grid, aes(load, prob), linewidth = 1) +
  labs(x = "Load", y = "Probability") +
  coord_cartesian(ylim = c(0, 1))
```

::: {.question}
At approximately what load does the fitted event probability cross 0.5?
:::

You can solve this directly from the fitted coefficients.

```{r}
b <- coef(fit_logit)
load_at_half <- -b[1] / b[2]
load_at_half
```

## 5. Predictions on three scales

```{r}
new_load <- data.frame(load = c(
  min(fastener$load),
  median(fastener$load),
  max(fastener$load)
))

data.frame(
  new_load,
  linear_predictor = predict(fit_logit, newdata = new_load, type = "link"),
  probability = predict(fit_logit, newdata = new_load, type = "response")
)
```

::: {.question}
Explain the distinction among probability, odds, and log-odds.
:::

## 6. Likelihood-based model comparison

Compare the logistic model with an intercept-only model.

```{r}
fit_null <- glm(
  cbind(y, m - y) ~ 1,
  family = binomial,
  data = fastener
)

anova(fit_null, fit_logit, test = "Chisq")
```

::: {.question}
What hypothesis is being tested by this likelihood-ratio comparison?
:::

## 7. Classification as a decision rule

A 0.5 probability threshold is common, but it is not universally optimal.

Expand the grouped counts only for this illustration.

```{r}
fastener_binary <- fastener |>
  rowwise() |>
  reframe(
    load = load,
    outcome = c(rep(1, y), rep(0, m - y))
  ) |>
  ungroup() |>
  mutate(
    prob = predict(fit_logit, newdata = data.frame(load = load),
                   type = "response"),
    class = if_else(prob >= 0.5, 1, 0)
  )

with(fastener_binary, table(observed = outcome, predicted = class))
mean(fastener_binary$outcome == fastener_binary$class)
```

::: {.question}
Why might a threshold other than 0.5 be preferable in an application where
false positives and false negatives have different consequences?
:::


## 8. Extension: individual-level logistic regression with Whickham data

The grouped `fastener` example is useful because it matches the current course
slides. The `Whickham` data provide a complementary example in which each row
is an individual and one predictor is categorical.

```{r}
names(Whickham)
head(Whickham)
table(Whickham$outcome, Whickham$smoker)
```

Create a binary response equal to 1 for an individual who was alive at the
follow-up.

```{r}
whickham <- Whickham |>
  mutate(
    isAlive = as.integer(outcome == "Alive"),
    smoker = factor(smoker)
  )
```

Fit a model using age and smoking status.

```{r}
fit_whickham <- glm(
  isAlive ~ age + smoker,
  family = binomial,
  data = whickham
)

summary(fit_whickham)
```

Display coefficient estimates as odds ratios.

```{r}
broom::tidy(fit_whickham, conf.int = TRUE, exponentiate = TRUE)
```

::: {.question}
1. Interpret the odds ratio associated with age.
2. Interpret the smoking-status odds ratio while holding age fixed.
3. Why can a marginal comparison between smokers and nonsmokers differ from
   the age-adjusted logistic-regression comparison?
:::

Visualize the fitted probabilities by smoking status.

```{r}
whickham_grid <- expand.grid(
  age = seq(min(whickham$age), max(whickham$age), length.out = 200),
  smoker = levels(whickham$smoker)
)

whickham_grid$prob <- predict(
  fit_whickham,
  newdata = whickham_grid,
  type = "response"
)

ggplot(whickham_grid, aes(age, prob, linetype = smoker)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Age",
    y = "Estimated probability of being alive",
    linetype = "Smoking status"
  ) +
  coord_cartesian(ylim = c(0, 1))
```

### Does the age effect differ by smoking status?

```{r}
fit_whickham_int <- glm(
  isAlive ~ age * smoker,
  family = binomial,
  data = whickham
)

anova(fit_whickham, fit_whickham_int, test = "Chisq")
```

::: {.question}
What does the likelihood-ratio test say about adding the age-by-smoking
interaction? Distinguish statistical evidence from substantive importance.
:::

::: {.callout-tip}
## Why use both logistic examples?
The `fastener` data illustrate **grouped binomial** logistic regression, while
`Whickham` illustrates **individual binary outcomes**, adjustment for a
categorical predictor, odds ratios, and possible interaction. Together they
show that the same logistic-regression framework handles both data structures.
:::

## Takeaway

Logistic regression models a linear relationship on the **log-odds** scale,
which guarantees fitted probabilities between 0 and 1. Coefficients are most
naturally interpreted through odds ratios, while classification requires an
additional decision threshold.
