---
title: "Lab 4: Added-Variable Plots"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    theme:
      - cosmo
      - lab-style.scss
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

Use added-variable (partial regression) plots to understand the contribution
of one predictor after adjusting for the remaining predictors.

```{r}
library(ggplot2)
library(dplyr)
library(car)
```

## 1. New York restaurant data

The Zagat data contain price and customer ratings for Italian restaurants in
New York City.

```{r}
nyc <- nyc_restaurants
head(nyc)
```

We focus on `Price`, `Food`, `Decor`, `Service`, and `East`.

```{r}
vars <- c("Price", "Food", "Decor", "Service", "East")
round(cor(nyc[vars]), 2)
pairs(nyc[vars])
```

::: {.question}
1. Which variables have the strongest marginal correlations with `Price`?
2. Which predictors are strongly correlated with one another?
:::

## 2. Fit the full model

```{r}
mfull <- lm(Price ~ Food + Decor + Service + East, data = nyc)
summary(mfull)
```

```{r}
car::vif(mfull)
```

::: {.question}
Why can a predictor have a reasonably strong correlation with `Price` but a
small or nonsignificant coefficient in the multiple regression model?
:::

## 3. Construct an added-variable plot by hand

To study `Service` after adjusting for `Food`, `Decor`, and `East`:

```{r}
m_y <- lm(Price ~ Food + Decor + East, data = nyc)
m_x <- lm(Service ~ Food + Decor + East, data = nyc)

av_service <- data.frame(
  y_resid = resid(m_y),
  x_resid = resid(m_x)
)

ggplot(av_service, aes(x_resid, y_resid)) +
  geom_point(size = 2.5, alpha = 0.65) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Service residual after adjusting for Food, Decor, East",
    y = "Price residual after adjusting for Food, Decor, East"
  )
```

The slope in this partial regression equals the coefficient of `Service` in
the full model.

```{r}
coef(lm(y_resid ~ x_resid, data = av_service))[2]
coef(mfull)["Service"]
```

::: {.question}
Explain in words why residualizing both variables isolates the partial
relationship between `Service` and `Price`.
:::

## 4. Use `car` for all added-variable plots

```{r}
car::avPlots(mfull, id = TRUE)
```

::: {.question}
Compare the added-variable plots. Which predictor contributes the clearest
additional linear information about price after controlling for the others?
Do any observations appear influential?
:::

## 5. Added-variable plots and diagnostics

```{r}
car::influencePlot(mfull, id = "noteworthy")
```

::: {.question}
An added-variable plot can reveal curvature, unusual cases, or changes in
spread associated with one predictor. What do you see here that is less
obvious in the ordinary residual-versus-fitted plot?
:::

## Takeaway

A marginal scatterplot asks, "How are \(X_j\) and \(Y\) related?" An
added-variable plot asks the more relevant multiple-regression question,
"How are \(X_j\) and \(Y\) related **after the other predictors have been
accounted for**?"
