---
title: "Lab 3: Transformations and Model Improvement"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    theme:
      - cosmo
      - lab-style.scss
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals


::: {.callout-note}
## Connection to model adequacy
Use transformations only after examining the original-scale relationship and
diagnostics. A useful transformation should improve the **structure of the
model** (linearity, variance stability, or error behavior) and remain
interpretable in the application.
:::


Recognize when a transformation may improve a regression model, compare
candidate transformations graphically, and use the Box-Cox profile as a
systematic guide.

```{r}
library(ggplot2)
library(dplyr)
library(MASS)
```

## 1. Electric utility data

The `electric_utility` object comes from the older course materials. Begin by
checking its variable names.

```{r}
names(electric_utility)
head(electric_utility)
```

For this lab, the first column is electricity use \(x\) and the second is peak
demand \(y\). We create local names so the analysis is robust to descriptive
column names in the course file.

```{r}
utility <- data.frame(
  x = electric_utility[[1]],
  y = electric_utility[[2]]
)
```

```{r}
ggplot(utility, aes(x, y)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
fit0 <- lm(y ~ x, data = utility)
par(mfrow = c(1, 2))
plot(fit0, which = 1)
plot(fit0, which = 2)
par(mfrow = c(1, 1))
```

::: {.question}
What evidence suggests that the original-scale model may be inadequate?
:::

## 2. Try common response transformations

```{r}
utility <- utility |>
  mutate(
    sqrt_y = sqrt(y),
    log_y = log(y)
  )
```

```{r}
ggplot(utility, aes(x, sqrt_y)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(y = expression(sqrt(y)))
```

```{r}
ggplot(utility, aes(x, log_y)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(y = "log(y)")
```

::: {.question}
Which transformation gives the most nearly linear relationship and the most
stable spread?
:::

## 3. Compare diagnostics

```{r}
fit_sqrt <- lm(sqrt(y) ~ x, data = utility)
fit_log  <- lm(log(y) ~ x, data = utility)

par(mfrow = c(2, 2))
plot(fit_sqrt, which = 1:2)
plot(fit_log,  which = 1:2)
par(mfrow = c(1, 1))
```

## 4. Box-Cox transformation

The Box-Cox profile provides a more systematic way to choose a power
transformation for a **positive** response.

```{r}
bc <- MASS::boxcox(fit0, lambda = seq(-2, 2, by = 0.05))
lambda_hat <- bc$x[which.max(bc$y)]
lambda_hat
```

::: {.question}
Is the estimated \(\lambda\) close to a familiar transformation such as
\(\lambda=1\) (none), \(1/2\) (square root), or \(0\) (log)?
:::

## 5. Windmill data: transforming the predictor

The older course slides also contain a windmill example. Here the transformed
predictor can be more useful than transforming the response.

```{r}
names(windmill)
head(windmill)
```

Use the first two columns as the original \(x\) and \(y\).

```{r}
wind <- data.frame(
  x = windmill[[1]],
  y = windmill[[2]]
)

ggplot(wind, aes(x, y)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE)
```

Compare a linear model in \(x\) with a model in \(1/x\).

```{r}
fit_x <- lm(y ~ x, data = wind)
fit_inv <- lm(y ~ I(1 / x), data = wind)

c(
  original_R2 = summary(fit_x)$r.squared,
  inverse_x_R2 = summary(fit_inv)$r.squared
)
```

```{r}
par(mfrow = c(1, 2))
plot(fit_x, which = 1)
plot(fit_inv, which = 1)
par(mfrow = c(1, 1))
```

::: {.question}
Which representation is easier to justify from the scatterplot and residual
plot? Remember that a high \(R^2\) alone is not enough to establish model
adequacy.
:::

## Takeaway

Transformations should be motivated by the scientific relationship and by
model diagnostics. They are not simply a device for maximizing \(R^2\).
