---
title: "Lab 1: Understanding Regression Output"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    css: lab-style.css
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

By the end of this lab, you should be able to connect the output from `lm()`
and `summary()` to the quantities developed in simple linear regression:
least-squares estimates, residual variation, \(R^2\), coefficient tests, and
the ANOVA decomposition.

We use the **rocket propellant** data from the earlier course slides. The
response is propellant **strength**, and the explanatory variable is propellant
**age**.

```{r}
library(ggplot2)
library(broom)

rocket_propellant
```

## 1. Explore the relationship

```{r}
ggplot(rocket_propellant, aes(x = age, y = strength)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age", y = "Strength")
```

::: {.question}
1. Describe the direction, form, and strength of the relationship.
2. Are there any observations that look unusual?
:::

## 2. Fit the model

```{r}
fit <- lm(strength ~ age, data = rocket_propellant)
summary(fit)
```

::: {.question}
Interpret the estimated intercept and slope in context. Is the intercept
scientifically meaningful over the observed range of the data?
:::

## 3. Reconstruct the least-squares estimates

Let \(x_i\) denote age and \(y_i\) denote strength. Compute

\[
S_{xx}=\sum (x_i-\bar{x})^2,\qquad
S_{xy}=\sum (x_i-\bar{x})(y_i-\bar{y}).
\]

```{r}
x <- rocket_propellant$age
y <- rocket_propellant$strength

xbar <- mean(x)
ybar <- mean(y)
Sxx <- sum((x - xbar)^2)
Sxy <- sum((x - xbar) * (y - ybar))

b1 <- Sxy / Sxx
b0 <- ybar - b1 * xbar

c(intercept = b0, slope = b1)
coef(fit)
```

::: {.question}
Verify that your calculations agree with `coef(fit)`.
:::

## 4. Sums of squares and \(R^2\)

```{r}
yhat <- fitted(fit)
e <- resid(fit)

SSE <- sum(e^2)
SST <- sum((y - ybar)^2)
SSR <- sum((yhat - ybar)^2)

c(SST = SST, SSR = SSR, SSE = SSE, check = SST - SSR - SSE)
```

```{r}
R2 <- SSR / SST
R2
summary(fit)$r.squared
```

::: {.question}
Interpret \(R^2\) in the context of propellant strength and age.
:::

## 5. Test the slope

For simple linear regression,

\[
SE(\hat\beta_1)=\frac{s}{\sqrt{S_{xx}}}, \qquad
s=\sqrt{\frac{SSE}{n-2}}.
\]

```{r}
n <- nrow(rocket_propellant)
s <- sqrt(SSE / (n - 2))
se_b1 <- s / sqrt(Sxx)
t_b1 <- b1 / se_b1
p_b1 <- 2 * pt(-abs(t_b1), df = n - 2)

c(s = s, SE_slope = se_b1, t = t_b1, p_value = p_b1)
```

```{r}
broom::tidy(fit, conf.int = TRUE)
```

::: {.question}
State the null and alternative hypotheses for the slope test and give a
conclusion in context.
:::

## 6. ANOVA table

```{r}
anova(fit)
```

::: {.question}
For a simple linear regression, compare the \(F\)-statistic in the ANOVA table
with the squared \(t\)-statistic for the slope. What do you notice?
:::

## 7. Prediction

```{r}
predict(fit,
        newdata = data.frame(age = c(5, 10, 15)),
        interval = "prediction",
        level = 0.95)
```

::: {.question}
Explain the difference between a confidence interval for the mean response and
a prediction interval for a new observation.
:::

## Takeaway

R output is not a separate procedure from the formulas: it is a compact
summary of the same least-squares calculations, uncertainty estimates, and
ANOVA decomposition developed in class.
