---
title: "Lab 2: Residual Diagnostics"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    theme:
      - cosmo
      - lab-style.scss
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

Use residual plots to diagnose violations of linearity, constant variance, and
normality, and recognize unusual observations.

```{r}
library(ggplot2)
library(dplyr)
library(broom)
```

## 1. Start with a well-behaved model

```{r}
set.seed(4780)

n <- 100
x <- runif(n, 0, 10)
good <- data.frame(
  x = x,
  y = 5 + 2 * x + rnorm(n, sd = 3)
)

fit_good <- lm(y ~ x, data = good)

augment(fit_good) |>
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point() +
  labs(x = "Fitted values", y = "Residuals")
```

```{r}
augment(fit_good) |>
  ggplot(aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line() +
  labs(x = "Theoretical quantiles", y = "Standardized residuals")
```

::: {.question}
What features of these plots would make you comfortable with the linear model?
:::

## 2. Nonlinearity

```{r}
nonlinear <- data.frame(
  x = x,
  y = 5 + 2 * x + 0.7 * x^2 + rnorm(n, sd = 3)
)

fit_nonlinear <- lm(y ~ x, data = nonlinear)

augment(fit_nonlinear) |>
  ggplot(aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point()
```

::: {.question}
What pattern indicates that the linear mean function is inadequate?
:::

## 3. Nonconstant variance

```{r}
hetero <- data.frame(
  x = x,
  y = 5 + 2 * x + rnorm(n, sd = 0.4 + 0.6 * x)
)

fit_hetero <- lm(y ~ x, data = hetero)

augment(fit_hetero) |>
  ggplot(aes(.fitted, .resid)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_point()
```

::: {.question}
Describe the change in residual spread as fitted values increase.
:::

## 4. Nonnormal errors

```{r}
nonnormal <- data.frame(
  x = x,
  y = 5 + 2 * x + (rchisq(n, df = 2) - 2)
)

fit_nonnormal <- lm(y ~ x, data = nonnormal)

augment(fit_nonnormal) |>
  ggplot(aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line()
```

::: {.question}
Where does the Q-Q plot depart from the reference line?
:::

## 5. A real-data diagnostic example

The **delivery time** data are used repeatedly in the course. Fit delivery
time using number of cases and distance.

```{r}
fit_delivery <- lm(time ~ cases + distance, data = delivery)
summary(fit_delivery)
```

Use the four standard diagnostic plots:

```{r}
par(mfrow = c(2, 2))
plot(fit_delivery)
par(mfrow = c(1, 1))
```

Then inspect several case diagnostics directly.

```{r}
delivery_diag <- augment(fit_delivery) |>
  mutate(
    observation = row_number(),
    leverage = hatvalues(fit_delivery),
    cooks_d = cooks.distance(fit_delivery),
    rstudent = rstudent(fit_delivery)
  )

delivery_diag |>
  select(observation, .fitted, .resid, rstudent, leverage, cooks_d) |>
  arrange(desc(cooks_d)) |>
  head(6)
```

::: {.question}
1. Which assumptions appear most questionable?
2. Which observations deserve closer investigation?
3. Does an unusual observation automatically imply that it should be deleted?
:::

## 6. Deleted residuals and PRESS

```{r}
press_resid <- resid(fit_delivery) / (1 - hatvalues(fit_delivery))
PRESS <- sum(press_resid^2)
PRESS
```

::: {.question}
Why are deleted residuals generally larger in magnitude than ordinary
residuals? How does PRESS connect residual diagnostics with prediction?
:::


## 7. Leverage and influence are different ideas

A point can have an unusual predictor value (high leverage) without having a
large residual, and a large residual does not necessarily imply high leverage.
Influence combines these features by asking how much the fitted model changes.

```{r}
n <- nrow(delivery)
p_coef <- length(coef(fit_delivery))

delivery_diag |>
  mutate(
    high_leverage = leverage > 2 * p_coef / n,
    large_studentized = abs(rstudent) > 2
  ) |>
  select(observation, rstudent, leverage, cooks_d,
         high_leverage, large_studentized) |>
  arrange(desc(cooks_d))
```

::: {.question}
Identify an observation that is unusual mainly because of leverage and one
that is unusual mainly because of its residual, if such observations exist.
Which observations are most influential according to Cook's distance?
:::

::: {.callout-tip}
## A modeling principle
Diagnostics identify observations that deserve investigation. They do **not**
provide an automatic rule for deleting data. First check data quality,
scientific context, and sensitivity of conclusions.
:::
