---
title: "Lab 6: Variable Selection and Prediction"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    css: lab-style.css
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

Compare best-subsets and stepwise procedures, evaluate candidate models using
adjusted \(R^2\), Mallows' \(C_p\), AIC, BIC, and PRESS/predicted \(R^2\), and
recognize why automated selection cannot replace model checking.

```{r}
library(leaps)
library(dplyr)
```

## 1. Hospital manpower data

We use the same data as the 2026 variable-selection slides.

```{r}
fit_full <- lm(y ~ ., data = manpower)
summary(fit_full)
```

## 2. Best-subsets regression

```{r}
best <- leaps::regsubsets(y ~ ., data = manpower, nvmax = 5)
best_summary <- summary(best)

best_table <- data.frame(
  size = seq_along(best_summary$rsq),
  R2 = best_summary$rsq,
  adj_R2 = best_summary$adjr2,
  Cp = best_summary$cp,
  BIC = best_summary$bic
)

best_table
```

```{r}
best_summary$which
```

::: {.question}
Which model size is favored by adjusted \(R^2\), \(C_p\), and BIC? Do the
criteria agree?
:::

## 3. A function for PRESS and predicted \(R^2\)

For a fitted linear model,

\[
\operatorname{PRESS}
 = \sum_{i=1}^n \left(\frac{e_i}{1-h_{ii}}\right)^2,
\qquad
R^2_{\mathrm{pred}}
 = 1-\frac{\operatorname{PRESS}}{SST}.
\]

```{r}
press_stats <- function(fit) {
  e <- resid(fit)
  h <- hatvalues(fit)
  press <- sum((e / (1 - h))^2)
  y <- model.response(model.frame(fit))
  sst <- sum((y - mean(y))^2)

  c(PRESS = press, R2_pred = 1 - press / sst)
}

press_stats(fit_full)
```

## 4. Compare the best model of each size

```{r}
which_mat <- best_summary$which

model_metrics <- lapply(seq_len(nrow(which_mat)), function(i) {
  selected <- names(which_mat[i, ])[which_mat[i, ]]
  selected <- setdiff(selected, "(Intercept)")
  form <- reformulate(selected, response = "y")
  fit <- lm(form, data = manpower)
  ps <- press_stats(fit)

  data.frame(
    size = length(selected),
    predictors = paste(selected, collapse = " + "),
    adj_R2 = summary(fit)$adj.r.squared,
    Cp = best_summary$cp[i],
    AIC = AIC(fit),
    BIC = BIC(fit),
    PRESS = unname(ps["PRESS"]),
    R2_pred = unname(ps["R2_pred"])
  )
})

bind_rows(model_metrics)
```

::: {.question}
Which model would you choose if prediction were the primary goal? Is it the
same model selected by BIC or \(C_p\)?
:::

## 5. Forward, backward, and bidirectional stepwise AIC

Base R's `step()` uses AIC by default.

```{r}
fit_null <- lm(y ~ 1, data = manpower)

forward_aic <- step(
  fit_null,
  scope = list(lower = fit_null, upper = fit_full),
  direction = "forward",
  trace = 0
)

backward_aic <- step(fit_full, direction = "backward", trace = 0)

both_aic <- step(
  fit_full,
  scope = list(lower = fit_null, upper = fit_full),
  direction = "both",
  trace = 0
)

formula(forward_aic)
formula(backward_aic)
formula(both_aic)
```

::: {.question}
Do all three search directions arrive at the same model? Why can greedy
search procedures depend on the path they take?
:::

## 6. Final model checking

Choose one candidate model and inspect it.

```{r}
final_fit <- both_aic
summary(final_fit)

par(mfrow = c(2, 2))
plot(final_fit)
par(mfrow = c(1, 1))

press_stats(final_fit)
```

::: {.question}
Would you report this model solely because an automated procedure selected it?
What additional checks and subject-matter considerations are needed?
:::

::: {.callout-tip}
## Important
Variable selection is a **model-building aid**, not a substitute for
scientific judgment, diagnostics, validation, or careful interpretation.
:::
