---
title: "Lab 5: Multicollinearity and VIF"
subtitle: "MATH 4780 / MSSC 5780 Regression Analysis"
author: "Mehdi Maadooliat"
format:
  html:
    css: lab-style.css
    toc: true
    toc-depth: 3
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r}
#| label: setup
#| include: FALSE
load("regression_datasets.rda")
options(digits = 3)
```

::: {.callout-note}
## Course data
This lab uses objects in `regression_datasets.rda`. Put the `.qmd` file and
`regression_datasets.rda` in the same folder before rendering or working
through the lab.
:::

## Learning goals

Diagnose multicollinearity using correlations and variance inflation factors
(VIFs), understand its effect on coefficient estimates, and compare nested
models.

```{r}
library(ggplot2)
library(car)
```

## 1. Hospital manpower data

This is the same hospital manpower dataset used in the 2026 multicollinearity
slides.

- \(y\): monthly man-hours
- \(x_1\): average daily patient load
- \(x_2\): monthly X-ray exposures
- \(x_3\): monthly occupied bed days
- \(x_4\): eligible population in the area / 1000
- \(x_5\): average length of patient stay in days

```{r}
manpower
pairs(manpower)
```

## 2. The full model

```{r}
fit_full <- lm(y ~ ., data = manpower)
summary(fit_full)
```

::: {.question}
Do any coefficient signs or significance tests seem surprising relative to
the marginal relationships in the pairs plot?
:::

## 3. Correlations among the predictors

```{r}
round(cor(manpower[-1]), 3)
```

::: {.question}
Which pairs of predictors appear most strongly correlated? Why is a
correlation matrix not sufficient to diagnose every form of
multicollinearity?
:::

## 4. Variance inflation factors

```{r}
vif_full <- car::vif(fit_full)
vif_full
```

Recall

\[
\operatorname{VIF}_j=\frac{1}{1-R_j^2},
\]

where \(R_j^2\) is obtained by regressing predictor \(X_j\) on all the other
predictors.

Verify one VIF manually.

```{r}
aux_x3 <- lm(x3 ~ x1 + x2 + x4 + x5, data = manpower)
1 / (1 - summary(aux_x3)$r.squared)
vif_full["x3"]
```

::: {.question}
Explain why a large \(R_j^2\) in the auxiliary regression inflates the
variance of \(\hat\beta_j\).
:::

## 5. Remove one redundant predictor

The 2026 slides investigate a model without `x3`.

```{r}
fit_no_x3 <- update(fit_full, . ~ . - x3)

summary(fit_no_x3)
car::vif(fit_no_x3)
```

Compare the models with a nested \(F\)-test.

```{r}
anova(fit_no_x3, fit_full)
```

::: {.question}
1. How do the coefficient estimates and standard errors change?
2. How do the VIFs change?
3. Is deleting `x3` justified by the nested \(F\)-test alone?
4. What substantive information might be lost by deleting a predictor?
:::

## 6. Prediction versus coefficient stability

```{r}
c(
  full_R2 = summary(fit_full)$r.squared,
  reduced_R2 = summary(fit_no_x3)$r.squared
)
```

::: {.question}
Why can a model have substantial multicollinearity and still fit the observed
responses well? Why is extrapolation particularly risky?
:::


## 7. A matrix view: condition number

VIFs diagnose predictor-specific inflation. A complementary global diagnostic
comes from the geometry of the standardized predictor matrix.

```{r}
X <- model.matrix(fit_full)[, -1, drop = FALSE]
Xz <- scale(X)

eigenvalues <- eigen(cor(Xz), symmetric = TRUE)$values
condition_index <- sqrt(max(eigenvalues) / eigenvalues)

data.frame(
  eigenvalue = eigenvalues,
  condition_index = condition_index
)
```

::: {.question}
Why can a very small eigenvalue indicate that some linear combination of the
predictors is nearly redundant?
:::

## Takeaway

Multicollinearity is primarily a problem of **unstable estimation and
interpretation**, not necessarily poor in-sample fit. VIF quantifies how much
the variance of an individual coefficient is inflated by linear dependence
among the predictors.
